---
title: 'Regression techniques with R data.table and h2o machine learning'
date: '`r Sys.Date()`'
author: "Bart Boerman"
output:
  html_document:
    theme: lumen
    toc: true
    number_sections: true
    font-family: Open Sans, sans-serif
    font-import: https://fonts.googleapis.com/css?family=Open+Sans
    code_folding: show
    highlight: tango
editor_options: 
  chunk_output_type: console
---
<style type="text/css">

body{ /* Normal  */
      font-size: 14px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1, .h1, h2, .h2, h3, .h3 {
    margin-top: 10.5px;
    margin-bottom: 10.5px;
}
h1.title {
  font-size: 28px;
  color: #7db956;
}
h1 { /* Header 1 */
  font-size: 28px;
  color: #3e4a52;
}
h2 { /* Header 2 */
    font-size: 18px;
  color: #3e4a52;
}
h3 { /* Header 3 */
  font-size: 14px;
  color: #3e4a52;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>
# Introduction

## Version

**Raw inital version**.

## Goal

Exploratory analyis and machine learning modal for predicting 
housing prices in competition ["House Prices: Advanced Regression Techniques"](https://www.kaggle.com/c/house-prices-advanced-regression-techniques). The aim is to predict house prices based on the provided data:

- **train.csv**,data for training our model. 
- **test.csv**, data used to see how well our model performs on unseen data.

In addition I want to gain and share some basic knowledge of

- data wrangling and analysis with data.table
- machine learning with h2o
- stacking with h2o

## Required libraries

- **knitr**, used to create this document 
- **data.table**, fast data wrangling with R
- **h2o**, machine learning algorithmes and more from h2o.ai
- **psych**, descriptive analytics, skewness and kurtosis
- **caret**, (near) zero variance
```{r setup, message=FALSE, warning=FALSE, paged.print=FALSE}
require(knitr) ## a general-purpose programming engine
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
require(data.table) ## fast data wrangling
require(h2o)        ## machine learning algorithmes
require(psych)      ## descriptive statistics, skewness and kurtosis
require(caret)      ## zero variance 
```

# Get data into R

```{r}
#### Fetch data
train.dt <- fread(input = "train.csv", 
                  sep = ",", 
                  nrows = -1,
                  header = T,
                  na.strings=c("NA","N/A","null"),
                  stringsAsFactors = F,
                  check.names = T,
                  strip.white = T,
                  blank.lines.skip = T,
                  data.table = T
) 
test.dt <- fread(input = "test.csv", 
                 sep = ",", 
                 nrows = -1,
                 header = T,
                 na.strings=c("NA","N/A","null"),
                 stringsAsFactors = F,
                 check.names = T,
                 strip.white = T,
                 blank.lines.skip = T,
                 data.table = T
) 
## Create one data set for feature engineering 
train.dt[, dataPartition:="train"]
test.dt[, SalePrice:=NA] 
test.dt[, dataPartition:="test"]
full.dt <- rbindlist(list(train.dt, test.dt), use.names = F, fill = F)
```

# Data dictionary

## Variables in raw data

```{r}
## Data types
variableTypes.df <- cbind(as.data.frame(names(full.dt)),as.data.frame(sapply(full.dt, class)))
names(variableTypes.df) <- c("variable","type")
kable(variableTypes.df, , row.names = F)
```

## Definitions

```{r}
## Numeric, square footage
variablesSquareFootage <- c(
  "LotFrontage", 		## Linear feet of street connected to property 
  "LotArea",    		## Lot size in square feet
  "MasVnrArea",  		## Masonry veneer area in square feet
  "BsmtFinSF1",		  ## Type 1 finished square feet	
  "BsmtFinSF2",		  ## Type 2 finished square feet
  "BsmtUnfSF",		  ## Unfinished square feet of basement area
  "TotalBsmtSF", 		## Total square feet of basement area
  "FirstFlrSF",		  ## First Floor square feet
  "SecondFlrSF",	  ## Second floor square feet
  "LowQualFinSF", 	## Low quality finished square feet (all floors)
  "GrLivArea", 		  ## Above grade (ground) living area square feet
  "GarageArea",     ## Size of garage in square feet
  "WoodDeckSF",     ## Wood deck area in square feet
  "OpenPorchSF",    ## Open porch area in square feet  
  "EnclosedPorch",  ## Enclosed porch area in square feet 
  "ThreeSsnPorch",  ## Three season porch area in square feet 
  "ScreenPorch",    ## Screen porch area in square feet
  "PoolArea" 		    ## Pool area in square feet
)
## Counts, a house has n of something
variablesCounts <- c(
  "BsmtFullBath",		## Basement full bathrooms
  "BsmtHalfBath",		## Basement half bathrooms
  "FullBath",			  ## Full bathrooms above grade
  "HalfBath",			  ## Half baths above grade
  "BedroomAbvGr",		## Bedrooms above grade (does NOT include basement bedrooms)
  "KitchenAbvGr",		## Kitchens above grade
  "TotRmsAbvGrd",		## Total rooms above grade (does not include bathrooms)
  "Fireplaces",		  ## Number of fireplaces
  "GarageCars"     	## Size of garage in car capacity
)
## Values
variablesValues <- c(
  "MiscVal",        ## $ Value of miscellaneous feature
  "SalePrice"       ## $ Price paid
)
## Factors
variablesFactor <- colnames(full.dt)[which(as.vector(full.dt[,sapply(full.dt, class)]) == "character")]
variablesFactor <- c(variablesFactor,  ## Add integers which are factors
                     "MSSubClass",     ## Identifies the type of dwelling involved in the sale
                     "OverallQual",    ## Rates the overall material and finish of the house
                     "OverallCond"     ## Rates the overall condition of the house
)
```

## Data conversion

```{r}
## In R first character can not be a number in variable names
setnames(full.dt, c("X1stFlrSF","X2ndFlrSF","X3SsnPorch"), c("FirstFlrSF","SecondFlrSF","ThreeSsnPorch"))
## Set columns to numeric
changeColType <- c(variablesSquareFootage, variablesCounts, variablesValues)
full.dt[,(changeColType):= lapply(.SD, as.numeric), .SDcols = changeColType]
## Set columns to factor
changeColType <- variablesFactor
full.dt[,(changeColType):= lapply(.SD, as.factor), .SDcols = changeColType]
```

# Descriptive statistics

## statisticts

```{r}
descStats <- describe(full.dt[, c(variablesSquareFootage,variablesValues), with = FALSE]) ## from psych package 
kable(descStats, digits = 2)
```

## Missing values

```{r}
countIsNA <- sapply(full.dt,function(x)sum(is.na(x)))
countIsNA.df <- data.frame(countIsNA)
countIsNA.df <- data.frame(variableName = row.names(countIsNA.df), countIsNA.df,row.names = NULL)
countIsNA.df <- countIsNA.df[countIsNA >0,]
kable(countIsNA.df)
```

## (Near) zero variance

```{r}
zeroVarianceVariables <- nearZeroVar(full.dt, names = TRUE, 
                                     freqCut = 19, uniqueCut = 10)
kable(zeroVarianceVariables)
```


## Skewness

```{r}
## skewness of numerical variables
skewedVariables <- sapply(full.dt[, c(variablesSquareFootage,variablesValues), with = FALSE],function(x){skew(x,na.rm=TRUE)}) ## from psych package
## keep only features that exceed a threshold for skewness
skewedVariables <- skewedVariables[skewedVariables > 0.75]
kable(skewedVariables, digits = 2, caption = "List of skewed variables")
```

## Correlation

To do.

# Feature engineering

## Set response variable

```{r}
response <- "SalePrice"
```

## Engineering

This stuff will be done soon.

## Remove zero variance variables

Consider engineering to capture information from these variables. On the "to do" list it goes.

```{r}
full.dt <- full.dt[, -c(zeroVarianceVariables), with = FALSE]
variablesSquareFootage <- setdiff(c(variablesSquareFootage), c(zeroVarianceVariables))
variablesValues      <- setdiff(c(variablesValues ), c(zeroVarianceVariables))

```

## Log transform skewed

```{r}
# Some items of .SDcols are not column names (or are NA)
# cols <- names(skewedVariables)
# full.dt[, (cols) := lapply(.SD, function(x) log(x)), .SDcols = cols]
```

## Scale numerical

```{r}
varScale <- setdiff(c(variablesSquareFootage, variablesValues), c(response)) ## Do not scale response
full.dt <- full.dt[ , (variablesSquareFootage) := lapply(.SD, scale), .SDcols = variablesSquareFootage]
```

## Select features

```{r}
features <- setdiff(names(full.dt), c(response, "Id","SalePrice","dataPartition")) 
```


